Tackling discrepancies in freshwater invertebrate trait databases and taxonomic resolution: Harmonising across continents and trait aggregation

### 2020-12-22 Paper/Draft_harmonization_paper.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{DRAFT: Harmonisation and trait aggregation paper }
\author{}%Stefan Kunz 
\date{}

@@

1. Invertebrate traits are expected to be suitable for large-scale analysis in freshwater ecology due to their consistent response to environmental gradients. In recent years, comprehensive invertebrate trait databases have been established at different scales (e.g. regions, continents). The wide availability of invertebrate trait data supports trait-based studies, especially on large scales. However, a number of data-related problems hamper the work with trait data. For example, for freshwater invertebrate traits standardised definitions are lacking and coding of traits varies across trait databases impeding trait data synthesis. Moreover, uncertainty remains how harmonising grouping features from different trait databases can influence trait-environment relationships. Another issue is that if the taxonomic resolution of observational taxonomic data is less precise than of trait databases, the traits in databases require aggregation before they can be linked to the observational data.
% Here: Describe aims of the study or redundant with part 2? Something like this: The aim of our study is to investigate these data-related issues by synthesizing trait data from different databases.

@@

3. Between the trait databases, grouping features varied in terms of their classification into traits, the scope of the grouping feature definitions, how traits are defined, and the coding to describe traits. Aggregated and assigned traits based on expert knowledge displayed a good match, specially approaches using the median. Grouping feature harmonisation and trait aggregation only slightly changed the results in the re-analysis compared to the original study, although averaging effects can occur due to harmonisation.

@@

In the last decades, freshwater ecologists have compiled comprehensive invertebrate trait databases for various continents (\cite{usseglio-polatera_biomonitoring_2000, schmidt-kloiber_www.freshwaterecology.info_2015, vieira_database_nodate, Philips_and_Smith_NZ_DB_2018, kefford_integrated_2020, tomanova_trophic_2006}). The availability of invertebrate trait data from different continents enables comparisons of trait variation and analyses of their relation to environmental factors across large scales. However, such analyses have been carried out mostly within continents, using information from one or two trait databases. For example, \citet{bonada_taxonomic_2007} compared trait composition for Mediterranean and temperate regions in Europe using traits from \citet{usseglio-polatera_biomonitoring_2000} (typically referred to as Tachet database); \citet{poff_developing_2010}) characterized trait composition across sites in the Western US using traits from \citet{poff_functional_2006}, and \citet{botwe_effects_2018} investigated the effect of salinity on invertebrate traits in different sites in South Australia using trait data from \citet{poff_functional_2006}) and \citet{schafer_trait_2011}. Analyses of invertebrate traits that synthesize information on invertebrate grouping features from more than two different continents are rare.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Tackling discrepancies in freshwater invertebrate trait databases and taxonomic resolution: Harmonising across continents and trait aggregation}
\author{}%Stefan Kunz 
\date{}

@@

1. Invertebrate traits are expected to be suitable for large-scale analysis in freshwater ecology due to their consistent response to environmental gradients. In recent years, comprehensive invertebrate trait databases have been established at different scales (e.g. regions, continents). The wide availability of invertebrate trait data supports trait-based studies, especially on large scales. However, a number of data-related problems hamper the work with trait data. For example, for freshwater invertebrate traits standardised definitions are lacking impeding trait data synthesis. Moreover, uncertainty remains how harmonising grouping features from different trait databases can influence trait-environment relationships. Another issue is that if the taxonomic resolution of observational taxonomic data is less precise than in trait databases, the traits in databases require aggregation before they can be linked to the observational data.
% Here: Describe aims of the study or redundant with part 2? Something like this: The aim of our study is to investigate these data-related issues by synthesizing trait data from different databases.

@@

3. Between the trait databases, grouping features differed in terms of their classification into traits and their scope leading to discrepancies how traits are defined. Also, the coding to describe traits varied between databases. Aggregated and assigned traits based on expert knowledge displayed a good match, specially approaches using the median. Grouping feature harmonisation and trait aggregation only slightly changed the results in the re-analysis compared to the original study, although averaging effects can occur due to harmonisation.

@@

In the last decades, freshwater ecologists have compiled comprehensive invertebrate trait databases for various continents (\cite{usseglio-polatera_biomonitoring_2000, schmidt-kloiber_www.freshwaterecology.info_2015, vieira_database_nodate, Philips_and_Smith_NZ_DB_2018, kefford_integrated_2020, tomanova_trophic_2006}). The availability of invertebrate trait data from different continents enables comparisons of trait variation and analyses of their relation to environmental factors across large scales. However, such analyses have been carried out mostly within continents, using information from one or two trait databases. For example, \citet{bonada_taxonomic_2007} compared trait composition for Mediterranean and temperate regions in Europe using traits from \citet{usseglio-polatera_biomonitoring_2000} (typically referred to as Tachet database); \citet{poff_developing_2010} characterized trait composition across sites in the Western US using traits from \citet{poff_functional_2006}, and \citet{botwe_effects_2018} investigated the effect of salinity on invertebrate traits in different sites in South Australia using trait data from \citet{poff_functional_2006}) and \citet{schafer_trait_2011}. Analyses of invertebrate traits that synthesize information on invertebrate grouping features from more than two different continents are rare.
